### 2025-04-04 latex/HP4_comm.tex
new files

%Look into referent of nirañjana in these texts. Judit: best to take it generally (like nirāmaya, etc.).
Verses 4.0*1–4.0*16 are omitted by the \textalpha\ group and are likely not to be original. The first additional verse resembles a \emph{maṅgala} verse that one might expect to see at the beginning of a text. The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samadhi}. Some manuscripts of the \textbeta, \textepsilon\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.29). The rest of the additional verses (4.0*5–4.0*16) are a motley collection on \emph{samādhi}, \emph{rājayoga}, the importance of the guru, dissolving the breath, \emph{suṣumṇā}, etc. In contrast to this, the \textalpha\ group begins with a cohesive discusson on absorption (4.1–3) that transitions to the gaze (4.4) and a brief discussion of \emph{śāmbhavī} and \emph{khecarī mudrā}s (4.5–4.8). The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.9 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses). The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.

Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.29).

@@

%%%%%%%%%%
\subsection*{4.0*2}
\begin{translation}[hp04_000_2]
So now I will teach the best way to \emph{samādhi}.It destroys death, has an easy method and brings about the bliss of Brahman.
\end{translation}

@@

@@

%%%%%%%%%%
\subsection*{4.0*7}
\begin{translation}[hp04_000_7]
Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation is established [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
\end{translation}

@@

\begin{philcomm}[hp04_000_7]
The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}. Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}. The compound \emph{muktisthitiḥ} (literally, `the establishment of liberation') has been adopted. 
%  attested by \etaTwo, 
% adopt muktisthitiḥ and rewrite note. 
%
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.0*11}
\begin{translation}[hp04_000_11]
When the breath is flowing in the central channel, and the mind enters the void, the knower of yoga destroys all actions.
\end{translation}
%?

\begin{sources}[hp04_000_11]
\end{sources}

@@

%%%%%%%%%%
\subsection*{4.0*12}
\begin{translation}[hp04_000_12]
Homage to you, the immortal one. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
\end{translation}

@@

%\begin{philcomm}[hp04_000_12]
%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
%

@@

%%%%%%%%%%
\subsection*{4.0*13}
\begin{translation}[hp04_000_13]
When equanimity has been obtained, and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises †even for the person whose hope is for that†.
\end{translation}
%

@@

%%%%%%%%%%
\subsection*{4.0*15}
\begin{translation}[hp04_000_15]
Both mercury and the mind are unsteady by nature. [When either] mercury is fixed [or] the mind is fixed, nothing in the world is impossible.
\end{translation}

@@

\begin{philcomm}[hp04_000_16]
The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).

@@

\begin{philcomm}[hp04_001_1]
The manuscripts attest two versions of the second line. It appears that the reading of \textbeta, \textepsilon\ and \texteta\ groups, which we have adopted, was changed to remove the first person pronoun. The modified reading is prevalent in the \textgamma\ and \textdelta\ manuscripts (\emph{manaḥprāṇalayānando nāpi kaś cit vibhidyate}).  
\end{philcomm}
%

@@

\begin{philcomm}[hp04_005]
See Birch 2013:286 for more parallel verses.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.8*1}
\begin{translation}[hp04_008_1]
[The yogi] should fix the pupils on a light and slightly raise the eyebrows. This is the way of the preliminary yoga, which immediately brings about the beyond-mind state.
\end{translation}
% unmanīkārakaḥ ?

\begin{sources}[hp04_008_1]
\emph{Amanaska} 1.8 (South Indian Recension)
\begin{versinnote}
\tl{netre jyotiṣi saṃyojya kiṃ cid unnamayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkārakaḥ kṣaṇāt ||\\+}
\tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!} 
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_008_1]
\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tāre jyotiṣi saṃyojya kiṃ cid uccālayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkaraṇaḥ kṣaṇāt ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_008_1]
This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7. 
\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*2}
\begin{translation}[hp04_008_2]
Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
\end{translation}

\begin{sources}[hp04_008_2]
\emph{Amanaska} 1.11 (South Indian Recension)
\begin{versinnote}
\tl{ke cid āgamajālena ke cin nigamasaṅkulaiḥ |\\+}
\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_008_2]
\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ke cid āgamajālena ke cin niyamasaṃkulāḥ |\\+}
\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_008_2]
In the source text \emph{tārakam} refers to  Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.

On why this verse is in greyscale, see the note to 4.8*1.    
\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*3}
\begin{translation}[hp04_008_3]
With eyes half open, mind steady, gaze placed at the tip of the nose and inner state motionless, [the yogi] should lead the moon and sun to dissolution. He attains the state of the highest reality, the supreme essence, whose form is light and which is devoid of anything external and shining intensely. What more is to be said here?
\end{translation}

\begin{sources}[hp04_008_3]
\emph{Candrāvalokana} 3
\begin{versinnote}
\tl{ardhodghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
\tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
\tl{tatvaṃ tatpadam eti yat tu paramaṃ vācyaṃ kim atrādhikaṃ ||\\+}
\tl{\var{sthiramanā ] 4340 4345 T00788, sphuramanā 7970}\\+}
\tl{\var{nāsāgradattekṣaṇaś ] 7970 4345 T00788, nāsāgradaṭhattakṣaṇaś 4340}\\+}
\tl{\var{tatvaṃ tatpadam eti yat tu] 4340 4345 T00788, tatvaṃ-n-tatpadam eti yat kṛ 7970}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_008_3]
\emph{Yogacintāmaṇi} ff. 24v–25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaḥ\\+}
\tl{candrārkāv api līnatām upanayen niḥspandabhāvottare |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ  \\+}
\tl{tatvaṃ tat padam eti vastu paramaṃ vācyaṃ kim atrādhikam ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_008_3]
This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze. In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups.
\end{philcomm}

\begin{metre}[hp04_008_3]
Śārdūlavikrīḍita 
\end{metre}

%%%%%%%%%%
\subsection*{4.8*4}
%\begin{translation}[hp04_008_4]
%That which enters the aperture into the underworld, which exists at the base of Meru,  the wise yogin say that is the truth, the source of [all] rivers. The essence of the body flows from the moon. Because of that, death arises for people. [This] good mudrā should block that [essence]. Otherwise, there is no bodily perfection.
%\end{translation}
% See 3.49

%\begin{sources}[hp04_008_4]
%\end{sources}

% \begin{testimonia}[hp04_008_4]
% \emph{Haṭharatnāvalī} 4.30
% \begin{versinnote}
% \tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti \\+}
% \tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
% \tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
% \end{versinnote}

% \emph{Yogacintāmaṇi} f.~74v (attr.~to the \emph{Haṭhapradīpikā})
% \begin{versinnote}
% \tl{pātāle yad viyati śikhare merumūle tad asti\\+}
% \tl{tattvaṃ caitat pravadati sudhīs sammukhe nimnagānām |\\+}
% \tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ\\+}
% \tl{taṃ badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
% \end{versinnote}

>>>>

%Look into referent of nirañjana in these texts. Judit: best to take it generally (like nirāmaya, etc.).
Verses 4.0*1–4.0*16 are omitted by the \textalpha\ group and are likely not to be original. The first additional verse resembles a \emph{maṅgala} verse that one might expect to see at the beginning of a text. The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samādhi}. Some manuscripts of the \textbeta, \textepsilon, \textzeta\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.32). The rest of the additional verses (4.0*5–4.0*16) are a motley collection on \emph{samādhi}, \emph{rājayoga}, the importance of the guru, dissolving the breath, \emph{suṣumṇā}, etc. In contrast to this, the \textalpha\ group begins with a cohesive discusson on absorption (4.1–3) that transitions to the gaze (4.4) and a brief discussion of \emph{śāmbhavī} and \emph{khecarī mudrā}s (4.5–4.8). The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.12 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses). The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.\lb

Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.32).\lb

@@

%%%%%%%%%%
\subsection*{4.0*2}
\begin{translation}[hp04_000_2]
So now I will teach the best way to \emph{samādhi}. It destroys death, has an easy method and brings about the bliss of Brahman.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.0*3}
\begin{translation}[hp04_000_3]
The sovereign yoga (\emph{rājayoga}), meditative absorption (\emph{samādhi}), the beyond mind state (\emph{unmanī}), the transmental state (\textit{manonmanī}), [the sovereign yoga of] the lineage of immortals (\emph{amaraugha}), dissolution [of mind] (\emph{laya}), the [ultimate] reality (\emph{tattva}), void and not void (\textit{śūnyāśūnya}), the highest state (\emph{para pada}), [\dots]
\end{translation}

\subsection*{4.0*4}
\begin{translation}[hp04_000_4]
[\dots] no-mind (\emph{amanaska}), non-duality (\emph{advaita}), without support (\emph{nirālamba}), pure (\emph{nirañjana}), liberation in life (\emph{jīvanmukti}), innate (\emph{sahaja}) and the fourth [state] (\emph{turya}) are synonyms.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.0*7}
\begin{translation}[hp04_000_7]
Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation becomes steady [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
\end{translation}

@@

\begin{philcomm}[hp04_000_7]
The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}. Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthirā}, \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}. We have adopted \emph{muktiḥ sthirā}, the reading of \textepsilon, an important group for the grey-scaled verses. 
%  attested by \etaTwo, 
% adopt muktisthitiḥ and rewrite note. 
% MD: The actual reading is muktiḥ sthirā.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.0*11}
\begin{translation}[hp04_000_11]
When the breath is flowing in the central channel, and the mind enters the void, the expert destroys all actions.
\end{translation}
%?

% \begin{sources}[hp04_000_11]
% \end{sources}

@@

%%%%%%%%%%
\subsection*{4.0*12}
\begin{translation}[hp04_000_12]
O Lord of the Immortals, homage to you. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
\end{translation}
%?? MD: Should we perhaps read 'amaraugha' instead of 'amareśa'? Considering 4.11*30, it is possible for inanimate things to become objects of worship. And it may not be a coincidence that the next verse is a quotation from the Amaraugha.

@@

%\begin{philcomm}[hp04_000_12]
%It appears that this verse, which is in manuscripts of the \textbeta\ and \textzeta\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
%

@@

%%%%%%%%%%
\subsection*{4.0*13}
\begin{translation}[hp04_000_13]
When equanimity has been obtained, and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises †even for the person whose hope is for that†.
\end{translation}
%?? MD: tadâśā jīvite'pi ca? "then there is also the prospect of staying alive"?

@@

%%%%%%%%%%
\subsection*{4.0*15}
\begin{translation}[hp04_000_15]
Both mercury and the mind are flighty by nature. When mercury is stabilised [or] the mind is stabilised, nothing in the world is impossible.
\end{translation}

@@

\begin{philcomm}[hp04_000_16]
The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).\lb

@@

\begin{philcomm}[hp04_001_1]
The manuscripts attest two versions of the second line. It appears that the reading of \textbeta, \textepsilon\ and \textzeta\ groups, which we have adopted, was changed to remove the first person pronoun. The modified reading is prevalent in the \textgamma\ and \textdelta\ manuscripts (\emph{manaḥprāṇalayānando nāpi kaś cit vibhidyate}).  
\end{philcomm}
%

@@

\begin{philcomm}[hp04_005]
See Birch 2013: 286 for more parallel verses.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.9}
\begin{translation}[hp04_009]
That which enters the aperture into the underworld, which exists at the base of Meru, the wise yogin say that is the truth, the source of [all] rivers. The essence of the body flows from the moon. Because of that, death arises for people. One should dam that [essence] with the earth that is the good \textit{mudrā}. Otherwise, there is no bodily perfection.
\end{translation}
% See 3.49*2

% The commentary has been moved from the third chapter. Should be modified.

\begin{testimonia}[hp04_009]
\emph{Haṭharatnāvalī} 4.30
\begin{versinnote}
\tl{pātāle yad viśati suṣiraṃ merūmūlaṃ tad asti\\+}
\tl{tattvaṃ caitad vadati sudhā tanmukhaṃ nimnagānām |\\+}
\tl{candrāt sāraṃ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
\tl{tad badhnīyāt sukharatimṛdur nānyathā kāryasiddhiḥ ||\\!}
\end{versinnote}
% after (ekaṃ sṛṣṭimayaṃ + mano yatra vilīyeta) and before (śarīraṃ tāvad eva)
%the context of this verse in the HR is rājayoga.

\emph{Yogacintāmaṇi} f.~75r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tat pātālād viyati śikhare merumūle tad asti \\+}
\tl{tattvaṃ caitat pravadati sudhīḥ saṃmukhe nimnagānām |\\+}
\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
\tl{tad badhnīyāt svakaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
\end{versinnote}
% after the verse (cittāyattaṃ nṛṇāṃ śukraṃ) and before (indhanāni yathā vahnis)
% context is khecarī (3rd chapter material)

\emph{Yuktabhavadeva} 7.220 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{pātāle yad viśati suṣiraṃ merumūle yad asti \\+}
\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇām \\+}
\tl{badhnīyāt tat sukaraṇam atho nānyathā kāyasiddhiḥ ||\\!}
\end{versinnote}
% context is khecarī, (3rd chapter material)
% after (ekaṃ sṛṣṭimayaṃ bījaṃ) and before (atha uḍḍiyānabandhaḥ)

\emph{Haṭhatattvakaumudī} 14.26 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{pātāle yad vitatasuṣiraṃ merumūle tad asmin \\+}
\tl{tadvac caitat pravadati sudhīs tanmukhaṃ nimnagānām |\\+}
\tl{candrāt sāraḥ sravati vapuṣas tena mṛtyur narāṇāṃ \\+}
\tl{taṃ badhnīyāt sukaraṇamṛdā nānyathā kāyasiddhiḥ ||\\!}
\end{versinnote}
% context is khecarī 3rd chapter material
% after (cumbantī yadi lambikā) and before (cittaṃ vicarati gagane)

\begin{philcomm}[hp04_009]
%Workshop notes
% Sven: the first line could refer to Kuṇḍalinī
% J5 has this verse here.
% mṛdā = using clay to make a dam?
% Also see Netratantra 7.32
% Could this verse be about mūlabandha? (reading pātāle and mūle)
%J5: yat prāleyam pihitasukhire merumūle yad astī — this seems helpful.
% PN has a nice-ish reading: yat prāleyaṃ pihita sukhiraṃ merumūrdhnaṃ. mūrdhni etc. is much better.
% J10 has merumūrdhni sthitaṃ yat, which is good.
% If going with prāleyam, adopt yat prāleyam pihitasuṣiraṃ merumūrdhny asti tathyaṃ (~K1 Grp2)
% ... tasmin tattvaṃ pravadati
%Haru: Perhaps, praleyam was corrupted to pātāle and then the rest of the line changed.

% α1,δ1 - in Ch. 4 only
% β1,β2 - after 3.39
% all the other mss - at 3.49*2
% among them α2,α3,gamma have it in both Ch. 3 and 4
On the position of these verses in the various recensions of the text, see the introduction (ref??).\lb

%This verse occurs in various places in the different recensions of the text. It is here in chapter four in the \textalpha\ and \textgamma\ groups (\alphaTwo\ and \alphaThree\ also have it in the third chapter at 3.49*2). It occurs after 3.39 in \betaOne\ and \betaTwo, and in both the third and fourth chapters in the \textgamma\ manuscripts. It must have entered the transmission of the \emph{Haṭhapradīpikā} at an early stage, and has moved around, perhaps because the name of the technique alluded to is not stated and the meaning of the first half of the verse is somewhat vague without its original context (we are yet to identify a source).

The first half of the verse has many variants among the witnesses. \alphaTwo\ has:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
\end{versinnote}
This version is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups. With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
\tl{\var{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.}\\!}
“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
\end{versinnote}

% Other manuscripts have something more similar to:
% \begin{versinnote}
% \tl{pātāle yad viśati suṣiraṃ merumūlaṃ tad asti \\+}
% \tl{tattvaṃ caitat pravadati sudhīs tan mukhaṃ nimnagānām\\!}
% \end{versinnote}

% MD: It is most likely that these and two other stanzas were here first and were moved to the 3rd chapter by the person who reorganised the HP. I would therefore suggest reactivating these stanzas here. Agreed.

\begin{metre}[hp04_009]
Mandākrāntā 
\end{metre}

%%%%%%%%%%
\subsection*{4.10}
\begin{translation}[hp04_010]
The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
\end{translation}

\begin{testimonia}[hp04_010]
\emph{Haṭharatnāvalī} 2.72 
\begin{versinnote}
\tl{atha viparītakaraṇī–\\+}
\tl{yat kiñ cit sravate candrād amṛtaṃ divyarūpi ca |\\+}
\tl{tatsarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~77v
\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{yat kiṃ cin sravate candrād amṛtaṃ divyarūpi ca |\\+}
\tl{tat sarvaṃ grasate sūryas tena piṇḍaṃ vināśi ca ||\\!}
\end{versinnote}

Cf.~\emph{Haṭhayogasaṃhitā} 38 (p. 26) % Gheraṇḍasaṃhitā 3.33
\begin{versinnote}
\tl{nābhimūle vaset sūryas tālumūle ca candramāḥ |\\+}
\tl{amṛtaṃ grasate sūryas tato mṛtyuvaśo naraḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_010]
%No source? %?? MD: VM 113-114 (Jim's working edition)
% nābhideśe vasaty eva bhāskaro dahanātmakaḥ |
% amṛtātmā sthito nityaṃ tālumadhye ca candramāḥ || 113 ||
% varṣaty adhomukhaś candro gṛhṇāty ūrdhvamukho raviḥ |
% jñātavyaṃ kāraṇaṃ tatra yena pīyūṣam āpyate || 114 ||
%(cited also in Jyotsnā 3.77)

%The term \emph{piṇḍa} is rarely neuter, hence the change to the masculine in the \emph{Jyotsnā}.

%are found in alpha in ch.4, but should they be there? They are very physical and not suited to ch.4. But here they are unusual because they are preamble and the practices already described do not have preamble. Perhaps drop them altogether. But then there is no explanation of how VK works.

%Supposedly these verses were originally here as in the alpha mss, first moved to the end of the Viparītakaraṇī section (as in the epsilon), and finally moved to the beginning of the same section (as in many other groups).
\end{philcomm}

%%%%%%%%%%
\subsection*{4.11}
\begin{translation}[hp04_011]
There is a divine bodily position for this, which blocks the mouth of the sun. It is to be known from the teaching of a guru and not through the countless interpretations of scriptures.
\end{translation}
% greyscaled

\begin{testimonia}[hp04_011]
Haṭharatnāvali 2.73 (on \emph{viparītakaraṇī})
\begin{versinnote}
\tl{tatrāsti divyaṃ karaṇaṃ sūryasya mukhabandhanam |\\+}
\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ || \\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~77v (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhabandhanam |\\+}
\tl{gurūpadeśato jñeyaṃ na tu śāstrārthakoṭibhiḥ ||\\!}
\end{versinnote}

%Yogakarṇikā 147c–148b
%\begin{versinnote}
%\tl{tat sarvaṃ grasate sūryas tena piṇḍo jarāyutaḥ || 147 ||\\+}
%\tl{tatrāsti karaṇaṃ divyaṃ sūryasya mukhavañcanam |\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_011]
%See the note on the previous verse.
%No source? Only Brahmānanda reads mukhavañcanam, which makes better sense. Judit; bandhanaṃ makes sense with grasate sūryas
\end{philcomm}

%%%%%%%%%%
\subsection*{4.11*1}
\begin{translation}[hp04_011_1]
[The yogi] should fix the pupils on a light and slightly raise the eyebrows. This is the way of the preliminary yoga, which immediately brings about the beyond-mind state.
\end{translation}
% unmanīkārakaḥ ?

\begin{sources}[hp04_011_1]
\emph{Amanaska} 1.8 (South Indian Recension)
\begin{versinnote}
\tl{netre jyotiṣi saṃyojya kiṃ cid unnamayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkārakaḥ kṣaṇāt ||\\+}
\tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!} 
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_011_1]
\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tāre jyotiṣi saṃyojya kiṃ cid uccālayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkaraṇaḥ kṣaṇāt ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_011_1]
This and the next verse (4.11*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7. 
\end{philcomm}

%%%%%%%%%%
\subsection*{4.11*2}
\begin{translation}[hp04_011_2]
Some are confused by a multitude of tantric texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
\end{translation}

\begin{sources}[hp04_011_2]
\emph{Amanaska} 1.11 (South Indian Recension)
\begin{versinnote}
\tl{ke cid āgamajālena ke cin nigamasaṅkulaiḥ |\\+}
\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_011_2]
\emph{Yogacintāmaṇi} f.~25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ke cid āgamajālena ke cin niyamasaṃkulāḥ |\\+}
\tl{ke cit tarkeṇa muhyanti naiva jānanti tārakam ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_011_2]
In the source text \emph{tārakam} refers to Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.\lb

On why this verse is in greyscale, see the note to 4.11*1.    
\end{philcomm}

%%%%%%%%%%
\subsection*{4.11*3}
\begin{translation}[hp04_011_3]
By leading the moon and sun to dissolution in a motionless state, the [yogi], his eyes half open, mind steady, and gaze placed at the tip of the nose, attains the supreme reality (\textit{vastu}), the state that is the highest principle (\textit{tattva}), 
% attains the state of the highest reality, the supreme essence (\textit{vastu}), 
whose form is light and which is devoid of anything external and shining intensely. What more is to be said here?
\end{translation}

\begin{sources}[hp04_011_3]
\emph{Candrāvalokana} 3
\begin{versinnote}
\tl{ardhodghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
\tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
\tl{tatvaṃ tatpadam eti yat tu paramaṃ vācyaṃ kim atrādhikaṃ ||\\+}
\tl{\var{sthiramanā ] 4340 4345 T00788, sphuramanā 7970}\\+}
\tl{\var{nāsāgradattekṣaṇaś ] 7970 4345 T00788, nāsāgradaṭhattakṣaṇaś 4340}\\+}
\tl{\var{tatvaṃ tatpadam eti yat tu] 4340 4345 T00788, tatvaṃ-n-tatpadam eti yat kṛ 7970}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_011_3]
\emph{Yogacintāmaṇi} ff. 24v–25r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ardhodghāṭitalocanaḥ sthiramanā nāsāgradattekṣaṇaḥ\\+}
\tl{candrārkāv api līnatām upanayen niḥspandabhāvottare |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ  \\+}
\tl{tatvaṃ tat padam eti vastu paramaṃ vācyaṃ kim atrādhikam ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_011_3]
This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze. In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups. The attentive reader will notice that we have decided against manuscripts that agree with the source text. The reason is based on our stemmatic considerations. Cumulative evidence supports the origin of the additional verses at the beginning of the fourth chapter in the hyparchetype of the \textepsilon\ manuscripts. We suspect these verses contaminated the \textbeta\ group's copies very early. To complicate matters, the scribe was familiar with the source text and corrected the readings accordingly.
\end{philcomm}

\begin{metre}[hp04_011_3]
Śārdūlavikrīḍita 
\end{metre}

%%%%%%%%%%
\subsection*{4.11*4}
\begin{translation}[hp04_011_4]
[The yogi] should not worship the \emph{liṅga} by day, nor should he worship it by night. He should worship the \emph{liṅga} constantly, by suppressing day and night.
% he should worship it day and night. Beta 2
\end{translation}

\begin{sources}[hp04_011_4]
\emph{Khecarīvidyā} 3.19
\begin{versinnote}
\tl{na divā pūjayel liṅgaṃ na rātrau ca maheśvari |\\+}
\tl{sarvadā pūjayel liṅgaṃ divārātrinirodhataḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_011_4]
\emph{Haṭhasaṅketacandrikā} f.~128v (attr.~Haṭhapradīpikā)
\begin{versinnote}
\tl{divā na pūjayel liṃga[ṃ] rātrau naiva prapūjayet [|]\\+}
\tl{satataṃ pūjayel liṃga[ṃ] divārātrau ca pūjayet [||]\\!}
\end{versinnote}
\end{testimonia}
